---
title: "Review ANOVA"
author: "Nzambuli Daniel"
date: "2024-05-15"
output: html_document
---

# Consider the following two sets of data: 

**Case 1: Equal Replications**

| Replicate | Treatment A | Treatment B | Treatment C |
|-----------|-------------|-------------|-------------|
| 1         | 23          | 42          | 47          |
| 2         | 36          | 26          | 43          |
| 3         | 31          | 47          | 43          |
| 4         | 33          | 34          | 39          |

**Case 1: Unequal Replications**

| Replicate | Treatment A | Treatment B | Treatment C | Treatment D |
|-----------|-------------|-------------|-------------|-------------|
| 1         | 2           | 1.7         | 2           | 2.1         |
| 2         | 2.2         | 1.9         | 2.4         | 2.2         |
| 3         | 1.8         | 1.5         | 2.7         | 2.2         |
| 4         | 2.3         |             | 2.5         | 1.9         |
| 5         | 1.7         |             | 2.4         |             |

## 1. Write the hypothesis to be tested

$h_0$ – the mean for each treatment has **no statistiacally significant difference**

$h_1$ – there is a **statistically significant** differenence in the mean of each treatment

## 2. Calculate the correction factor, Total SS, Treatment SS and Error SS

$$
experiment\ units = repetitions * treatments\\
n = r * t
$$

```{r}
equal_reps = data.frame(replicate = seq(1,4),
                        treat_A = c(23, 36, 31, 33),
                        treat_B = c(42, 26, 47, 34),
                        treat_C = c(47, 43, 43, 39)
)
print("Equal Repetition")
equal_reps
unequal_reps = data.frame(replicate = seq(1, 5),
                          treat_A = c(2, 2.2, 1.8, 2.3, 1.7),
                          treat_B = c(1.7, 1.9, 1.5, NA, NA),
                          treat_C = c(2, 2.4, 2.7, 2.5, 2.4),
                          treat_D = c(2.1, 2.2, 2.2, 1.9, NA))
print("Unequal Repetition")
unequal_reps
```

$$
correction\ factor = \frac{grand\ total ^ 2}{observations}\\
= \frac{(\sum^{treatments[A, B, C]}_{i=A}a \in i) ^ 2}{4 * 3}
$$

**Correction Factor for Equal Replicates**

```{r}
equal_n = nrow(equal_reps) * ncol(equal_reps[-1])
equal_exp_total = sum(equal_reps[-1])
equal_corr_f = (equal_exp_total^ 2)/ equal_n
cat("The correction factor of equal replicates is:\n", equal_corr_f, "\n")
```

**Unequal Replicates**

Remove the na before calculation

```{r}
library(dplyr)
library(tidyr)
```

```{r}
unequal_n =  sum(!is.na(unequal_reps[-1]))
unequal_exp_total = sum(unequal_reps[-1], na.rm = TRUE)
unequal_corr_f = (unequal_exp_total^ 2)/ unequal_n
cat("The correction factor of equal replicates is:\n", unequal_corr_f, "\n")
```

**Total Sum of squares equal replicates**

```{r}
equal_tss = sum(equal_reps[-1]^ 2) - equal_corr_f
cat("The equal replicates total sum of squares is:\n", equal_tss, "\n")
```

**Unequal replicates**

```{r}
unequal_tss = sum(unequal_reps[-1]^ 2, na.rm = TRUE) - unequal_corr_f
cat("The unequal replicates total sum of squares is:\n", unequal_tss, "\n")
```

**Treatment sum of squares equal replicates**

```{r}
equal_tts = colSums(equal_reps[-1])
equal_ttss = sum(equal_tts^2)/ nrow(equal_reps) - equal_corr_f
cat("The equal replicates treatment sum of squares is:\n", equal_ttss, "\n")
```

**unequal replicates**

```{r}
unequal_tts = colSums(unequal_reps[-1], na.rm = T)
unequal_ttss = sum(unequal_tts^2)/ colSums(!is.na(unequal_reps)) - unequal_corr_f
cat("The unequal replicates treatment sum of squares is:\n", unequal_ttss, "\n")
```

**Error sum of squares equal replicates**

```{r}
cat("The equal replicates error sum of squares is:\n", equal_tss - equal_ttss, "\n")
```

**Unequal replicates**

```{r}
cat("The equal replicates error sum of squares is:\n", unequal_tss - unequal_ttss, "\n")
```

## 3. Generate the ANOVA table

**Equal replicates**

```{r}
equal_long = equal_reps %>% gather(key = "treatment", value = "value", - replicate) # using every column except the replicate column
equal_long
```

**The anova table**

```{r}
anova_equal_reps = aov(value~treatment, data = equal_long)
summary(anova_equal_reps)
```

**Unequal replicates**

```{r}
# there are na in the unequal replicate data 
# using filter can help ~i hope 
unequal_long = unequal_reps %>% gather(key = "treatment", value = "value", - replicate) %>% filter(!is.na(value)) # using every column except the replicate column
unequal_long
```

**The anova table**

```{r}
anova_unequal_reps = aov(value~treatment, data = unequal_long)
summary(anova_unequal_reps)
```

## 4. At 5% and 10% levels of significance, test the hypothesis and make conclusions

Display the p-values

```{r}
equal_pval = summary(anova_equal_reps)[[1]][["Pr(>F)"]][1]
unequal_pval = summary(anova_unequal_reps)[[1]][["Pr(>F)"]][1]

hypothesis_conclusion = function(sig_lev, pval, case){
  cat("\nAt", round(pval, 4), "% significance level:\n")
  if(pval < sig_lev){
     cat("for case", case,"at significance level", sig_lev, "Reject the null hypothesis.\n\tConclude that there is a statistically significance difference in the treatment means")
  }else{
    cat("for case", case,"at significance level", sig_lev, "Fail to reject the null hypothesis.\n\tConclude that there is no statistically significance difference in the treatment means")
  }
}
```

### conclusions

**equal replica**

```{r}
hypothesis_conclusion(0.05, equal_pval, "equal replicates")
```

```{r}
hypothesis_conclusion(0.1, equal_pval, "equal replicates")
```

**unequal replica**

```{r}
hypothesis_conclusion(0.05, unequal_pval, "unequal replicates")
```

```{r}
hypothesis_conclusion(0.1, unequal_pval, "unequal replicates")
```
