---
output:
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## setup framework for rows, columns and treatments

```{r}

rows<- c(rep("r1",1), rep("r2",1), rep("r3",1), rep("r4",1), rep("r5",1))
cols<- c(rep("cA",5), rep("cB",5), rep("cC",5), rep("cD",5), rep("cE",5))
treatments<- c("A","C","D","E","B", "D","B","C","A","E", "E","A","B","C","D",
               "B","E","A","D","C", "C","D","E","B","A")


```


## assign data to variable "yield"

```{r}

yield<- c(7.4,11.8,10.1,8.8,11.8,8.9,6.5,17.9,10.1,8.8,5.8,8.7,9.0,15.7,14.3,
          12.0,7.6,8.5,11.1,18.4,14.3,7.9,7.1,7.4,10.1)
y<- data.frame(cols,rows,treatments, yield)
matrix(y$treatments, 5,5)
matrix(y$yield, 5,5)

```

## apply a linear model then compute the analysis of variance table

```{r}

yfit <- lm(yield ~ rows+cols+treatments, y)
anova(yfit)

```


## Combined Analysis Across Latin Squares

### Input the Data for Three 3x3 Latin Squares

```{r}
# Data for three Latin squares
square1 <- matrix(c(41, 25, 15, 20, 32, 24, 22, 12, 21), nrow=3, byrow=TRUE)
square2 <- matrix(c(27, 28, 3, 4, 17, 9, 22, 4, 17), nrow=3, byrow=TRUE)
square3 <- matrix(c(43, 27, 17, 22, 34, 26, 24, 14, 23), nrow=3, byrow=TRUE)

# Convert the data to a dataframe
df1 <- data.frame(row=rep(1:3, each=3), col=rep(1:3, times=3), 
                  trt=factor(c('B', 'C', 'A', 'A', 'B', 'C', 'C', 'A', 'B')),
                  yield=c(41, 25, 15, 20, 32, 24, 22, 12, 21))
df2 <- data.frame(row=rep(1:3, each=3), col=rep(1:3, times=3), 
                  trt=factor(c('C', 'B', 'A', 'A', 'C', 'B', 'B', 'A', 'C')),
                  yield=c(27, 28, 3, 4, 17, 9, 22, 4, 17))
df3 <- data.frame(row=rep(1:3, each=3), col=rep(1:3, times=3), 
                  trt=factor(c('B', 'C', 'A', 'A', 'B', 'C', 'C', 'A', 'B')),
                  yield=c(43, 27, 17, 22, 34, 26, 24, 14, 23))
```

### Perform Bartlett’s Chi-square Test for Homogeneity of Variance

```{r}
# Error sums of squares
error_ss1 <- 21.56
error_ss2 <- 14.89
error_ss3 <- 21.56

# Error degrees of freedom
error_df <- 2

# Error mean squares
error_ms1 <- error_ss1 / error_df
error_ms2 <- error_ss2 / error_df
error_ms3 <- error_ss3 / error_df

# Log of each error MS
log_error_ms1 <- log(error_ms1)
log_error_ms2 <- log(error_ms2)
log_error_ms3 <- log(error_ms3)

# Pooled Error MS
pooled_error_ms <- (error_ss1 + error_ss2 + error_ss3) / (3 * error_df)

# Bartlett's Chi-square test
chisq_value <- (3 * (log(pooled_error_ms) - (log_error_ms1 + log_error_ms2 + log_error_ms3) / 3)) * (3 * error_df)
p_value <- pchisq(chisq_value, df=2, lower.tail=FALSE)

# Print results
chisq_value
p_value
```

### Calculate Treatment Totals, Correction Factor, and Various Sums of Squares

```{r}
# Treatment totals
treatment_totals <- data.frame(
  treatment = c('A', 'B', 'C'),
  square1 = c(47, 94, 71),
  square2 = c(11, 59, 61),
  square3 = c(53, 100, 77)
)
treatment_totals$total <- rowSums(treatment_totals[, -1])

# Correction Factor (CF)
total_yield <- sum(treatment_totals$total)
correction_factor <- (total_yield^2) / (3 * 3 * 3)

# Total SS
total_ss <- sum((c(square1, square2, square3))^2) - correction_factor

# Square SS
square_ss <- sum(c(212, 131, 230)^2 / (3 * 3)) - correction_factor

# Row(Square) SS (Additive across squares)
row_ss <- sum(c(126.89, 130.89, 126.89))

# Column(Square) SS (Additive across squares)
column_ss <- sum(c(89.55, 110.22, 89.55))

# Treatment SS
treatment_ss <- sum(treatment_totals$total^2 / (3 * 3)) - correction_factor

# Square x Treatment SS
sq_x_trt_ss <- total_ss - square_ss - treatment_ss

# Error SS (Additive across squares)
error_ss <- sum(c(21.56, 14.89, 21.56))

# ANOVA table
anova_table <- data.frame(
  SOV = c('Square', 'Row(Sq)', 'Column(Sq)', 'Trt', 'Sq X Trt', 'Error', 'Total'),
  Df = c(2, 6, 6, 2, 4, 6, 26),
  SS = c(square_ss, row_ss, column_ss, treatment_ss, sq_x_trt_ss, error_ss, total_ss)
)
anova_table$MS <- anova_table$SS / anova_table$Df
anova_table$F <- c(NA, NA, NA, anova_table$MS[4] / anova_table$MS[6], anova_table$MS[5] / anova_table$MS[6], NA, NA)

# Print ANOVA table
anova_table
```

### Calculate the Least Significant Difference (LSD) for Treatments

```{r}
# LSD for treatment
alpha <- 0.05
t_critical <- qt(1 - alpha / 2, df = anova_table$Df[6])
lsd_trt <- t_critical * sqrt(2 * anova_table$MS[6] / (3 * 3))

# Print LSD
lsd_trt
```
## Importing data

```{r}

library(readxl)
data = read_excel(path = "data-LS Design.xlsx",
                col_names = TRUE)

head(data)
str(data)

```

## changing variables structure into factors


```{r}
data$Row <- as.factor(data$Row)
data$Column <- as.factor(data$Column)
data$Varieties <- as.factor(data$Varieties)
str(data)
attach(data)
```

## Applying analysis of variance model
```{r}

model <- lm(Yield ~ Row+Column+Varieties)
anova(model)

```


## Mean comparison test

```{r}

library(agricolae)
# LSD test
LSD.test(y = model,
         trt = "Varieties",
         DFerror = model$df.residual,
         MSerror = deviance(model)/model$df.residual,
         alpha = 0.05,
         group = TRUE,
         console = TRUE)


# SNK test
SNK.test(y = Yield,
         trt = Varieties,
         DFerror = model$df.residual,
         MSerror = deviance(model)/model$df.residual,
         alpha = 0.05,
         group = TRUE,
         console = TRUE)

```
